---
title: "Species Distribution Modelling"
---

```{r load_libraries, message=FALSE, warning=FALSE}
# Load libraries
library(tidyverse)
library(lubridate)
library(caret)
library(raster)
library(reshape2)
library(viridis)

# Clear environment and set working directory
rm(list = ls())  
```

We can use the Global Biodiversity Information Facility (GBIF) to gather presence records for the migratory hoverfly *Volucella zonaria*, also known as the hornet mimic hoverfly. In Great Britain, it was only known from two specimens prior to 1940, so was regarded as rare. Since then, it has become increasingly widespread in many parts of the South and South East England, often in association with parks and gardens, where adults are usually seen visiting flowers. Elsewhere in England, only a few scattered records exist.

```{r get_and_clean_data}

# load the data from a query to GBIF, for V. Zonaria
gbif_response <- read_rds("~/Documents/GitHub/ComputationalEcology/data_analysis_files/clim_data/gbif_volucella_zonaria.rds")

# Create presence records across decades
zonaria_clean <- gbif_response$data %>%
  # get decade of record from eventDate
  mutate(decade = eventDate %>% 
           ymd_hms() %>% 
           round_date("10y") %>% 
           year() %>% 
           as.numeric()) %>%
  rename(y = decimalLatitude,
         x = decimalLongitude) %>%
  # clean data using metadata filters
  filter(
    # only records with no issues
    issues == "" &
    # let's take data from 2000 to 2020
    decade %in% c(2010,2020)) %>%
  # retain only relevant variables
  dplyr::select(x, y, decade) %>% arrange(decade)
```

# Covariates

Now we need to grab some environmental data. For this vignette we will use the worldclim database as this is directly available via the `raster` package. We can download 19 bioclimatic variables at 10' resolution. They are coded as follows:

* BIO1 = Annual Mean Temperature
* BIO2 = Mean Diurnal Range (Mean of monthly (max temp - min temp))
* BIO3 = Isothermality (BIO2/BIO7) (* 100)
* BIO4 = Temperature Seasonality (standard deviation *100)
* BIO5 = Max Temperature of Warmest Month
* BIO6 = Min Temperature of Coldest Month
* BIO7 = Temperature Annual Range (BIO5-BIO6)
* BIO8 = Mean Temperature of Wettest Quarter
* BIO9 = Mean Temperature of Driest Quarter
* BIO10 = Mean Temperature of Warmest Quarter
* BIO11 = Mean Temperature of Coldest Quarter
* BIO12 = Annual Precipitation
* BIO13 = Precipitation of Wettest Month
* BIO14 = Precipitation of Driest Month
* BIO15 = Precipitation Seasonality (Coefficient of Variation)
* BIO16 = Precipitation of Wettest Quarter
* BIO17 = Precipitation of Driest Quarter
* BIO18 = Precipitation of Warmest Quarter
* BIO19 = Precipitation of Coldest Quarter

We'll also to use landcover classification for the UK (from 2015).


Let's download data for the current bioclim and UK lancover classification rasters:

```{r environmental_data, message=FALSE, warning=FALSE}
# Load UK background mask
bg <- raster("data_analysis_files/UK_bg.grd")

# Reproject to lat/lon
bg <- projectRaster(bg, crs= "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs")

# UK extent in lon/lat coordinates
ext_uk <- c(-12, 3, 48, 62)

# Get climate data for current date
filepath = "~/Documents/GitHub/ComputationalEcology/data_analysis_files/clim_data/"
bio_curr <- getData('worldclim', var='bio', download=F, lon=-5, lat=55, res=5, path=filepath)

# Crop and reproject current climate
bio_curr <- crop(bio_curr, ext_uk)
bio_curr <- projectRaster(bio_curr, bg)
bio_curr <- resample(bio_curr, bg)
bio_curr <- mask(bio_curr, bg)

# Future climate data
bio_fut <- getData('CMIP5', var='bio', download=F, res=5, rcp=45, model='MC', year=50, path=filepath)
bio_fut <- crop(bio_fut, ext_uk)

# Crop and reproject future climate
bio_fut <- crop(bio_fut, ext_uk)
bio_fut <- projectRaster(bio_fut, bg)
bio_fut <- resample(bio_fut, bg)
bio_fut <- mask(bio_fut, bg)

# Make variable names the same for both datasets
names(bio_fut) <- names(bio_curr)

# Load UK Landcover percentage classification at 1km
uk_lcm <- stack("data_analysis_files/lcm/LCM2015_GB_1km_percent_cover_target_class_WGS84.tif")

# Join uk landcover and current bioclim rasters
curr_clim_resampled <- resample(uk_lcm, bio_curr)
curr_clim_stacked <- stack(curr_clim_resampled, bio_curr)

# Join uk landcover and future bioclim rasters
fut_clim_resampled <- resample(uk_lcm, bio_fut)
fut_clim_stacked <- stack(fut_clim_resampled, bio_fut)


```

Let's plot our GBIF observations for V. zoanria
```{r plot_obs, message=FALSE, warning=FALSE}

bg_df <- as(bg, "SpatialPixelsDataFrame") %>% 
  as.data.frame()

colnames(bg_df) <- c("value", "x", "y")

ggplot() +
  geom_tile(data= bg_df, aes(x=x, y=y)) +
  coord_equal() +
  geom_point(data=zonaria_clean, 
             aes(x=x, y=y), 
             colour ="red", 
             size = 0.5) +
  theme_bw()

```

## Bioclim data

We can plot the current bioclim data - all 19 covariates:

```{r plot_uk_bioclim, fig.asp=4, fig.align='center', message=FALSE, warning=FALSE}
# Plot the bioclim covariates - we'll use ggplot, as raster plot function not as nice output

# Function to generate individual plots
plot_bioclim <- function(df, name) {
  ggplot(data = df, aes(x=x, y=y, fill = value)) + 
  geom_tile() +
  coord_equal() +
  scale_fill_viridis(name = name) +
  theme(
    plot.margin = rep(unit(0,"null"),4),
    panel.spacing = unit(0,"null"),
    legend.key.width = unit(0.01,"null")
    ) +
  theme_bw()
}

# extract all bioclim values by xy
nested_bioclim_plots <- data.frame(rasterToPoints(bio_curr)) %>%
  # Create single gouping variable by which to facet chart
  melt(id = c("x","y")) %>%
  # Nest data by bioclim variable
  nest(-variable) %>%
  # Create individual ggplots
  mutate(plots = map2(.x = data, 
                      .y = variable, 
                      .f = plot_bioclim))

# Render plots
library(gridExtra)
margin = theme(plot.margin = rep(unit(0,"null"),4))
gridExtra::grid.arrange(grobs = nested_bioclim_plots$plots,
                        top="Current UK Bioclim variables ",
                        ncol = 2,
                        margin
                        )
```

For each bioclim and uk landcover variable, we want to extract its value for the location of each V. zonaria observation. We also need to create pesudo-absence data - this represents locations where no observations were made. The covariates then need to be joined to the location of each location.

```{r extraxt_covariates}

df_presence <- zonaria_clean[1:2] %>%
  # Extract bioclim + lcm variables using long/lat
  raster::extract(curr_clim_stacked, .) %>%
  # Add extracted variables to long/lat
  cbind(zonaria_clean) %>%
  # Mark observation data as presence records
  mutate(presence = 1) %>%
  # Drop decade, we dont need it
  dplyr::select(-decade)
  
# Function to pick raster cells at random from bioclim layers
raster_random_sample <- function(r_covars, n_samples, xy_coords) {
  raster::sampleRandom(
    x = r_covars, 
    size = n_samples,
    na.rm = TRUE,
    xy = xy_coords) 
  }

# Sample background points (psedo absent?) from covariate raster 
df_absence <- raster_random_sample(curr_clim_stacked, 500, T) %>%
  as.data.frame() %>%
  # Mark as background data
  mutate(presence = 0)

# Create single dataframe for presence and absence data
df_zonaria <- rbind(df_presence, df_absence)
```

# Variable selection

In order to avoid collinearity we should remove covariates with absolute correlation of 0.8 or higher. We can achieve this by calculating a correlation matrix.

```{r corr_mat}
# Generate correlation matrix with covariates as correlates
cor_mat <- cor(df_zonaria %>% dplyr::select(-x,-y,-presence), method="spearman")
# Subset covariates that are highly correlated
highlyCorrelated <- findCorrelation(cor_mat, cutoff=0.75)
# Generate column names to remove
cols_to_remove <- names(df_zonaria[highlyCorrelated])
# Remove highly correlated variables from df
df_zonaria <- df_zonaria %>% dplyr::select(-cols_to_remove)
```

Now we have joined our environmental and observation data together into one dataset, we can begin to build a species distribution model. First we need to create a categorical response variable as we want to model presence v absence in space:

```{r pre_processing, warning=FALSE}
# last pre-processing step
df_modelling <- df_zonaria %>%
  # caret requires a factorial response variable for classification
  mutate(presence = case_when(
    presence == 1 ~ "presence",
    presence == 0 ~ "absence") %>%
    as.factor()) %>%
  # drop all observations with NA variables
  na.omit()
```

# Covariate importance

Before we fit the a machine learning model, we should look at the variabele importance. We can do this using the `caret` package, which has a function to evaluate covariate importance.

```{r feature_selection, warning=FALSE}
# Use Recursive Feature Elimintation (RFE) to determine significant covariates 
set.seed(100)

# Select the grouping of variables to test
subsets <- c(1:5, 10, 15, 20, length(df_modelling))

# Define the control using a random forest selction function
ctrl <- rfeControl(functions = rfFuncs,
                   method = "repeatedcv",
                   repeats = 3,
                   verbose = FALSE)

results <- rfe(x=df_modelling %>% dplyr::select(-c("x","y","presence")), 
               y=df_modelling$presence,
               sizes = subsets,
               rfeControl = ctrl)

# Print summart of results
print(results)
# Plot the results
plot(results, type=c("g", "o"))
```

# Fit a Random Forest ML model

Looks like we can get most of the way with the top five variables, so let's use the first five most significant variables in our model. We'll fit the model using a random forest.

```{r model_fit, message=FALSE, warning=FALSE, fig.align='centre'}
library(randomForest)

# take the top five variables from the feature importance
opt_vars <- results$optVariables[1:5]

# Recut the modelling data
df_modelling <- cbind(df_modelling[opt_vars], df_modelling %>% dplyr::select(x,y,presence))

# Use five fold cross validation, repeated twice
tunecontrol <- trainControl(
  method = 'repeatedcv',
  number = 5,
  repeats = 2,
  search = "grid")

# Create tuning parameters as grid
tunegrid <- expand.grid(.mtry=c(3:9))

# for reproducibility
set.seed(12345)

# actual model build
model_fit <- train(
  presence ~ .,
  data = df_modelling,
  method = "rf",
  metric = "Accuracy",
  tuneGrid = tunegrid,
  trControl = tunecontrol)

print(model_fit)
plot(model_fit)
```

# Make a prediction

Now we can now make a prediction for the current species distribution range for V. zonaria, using the bioclim covariates.

```{r predict_and_plot, message=FALSE, warning=FALSE, fig.align='centre', fig.width=9}
# Generate species distribution df given raster of environmental variables and sdm
predict_dist <- function(
  # Raster containing all covariates
  r_covar, 
  # Model fit
  sdm)
  # Subset of covariates required by model fit
  #opt_covars)
  {
     # Generate distribution data from raster to run model on
     dist_dat <- data.frame(rasterToPoints(r_covar)) %>% 
       dplyr::select(opt_vars,x,y) %>%
       na.omit()
     
     # Set seed for repeatable results
     set.seed(123)
     # Create a probability of presence/absence raster
     r_pred <- rasterFromXYZ(
       cbind(
       # Coordinates to make prediction from
       dist_dat %>% dplyr::select(x,y),
       # Predict with RF species distribution model 
       predict(model_fit, 
               # Use covariates to run model against
               dist_dat, 
               # Type of response variable
               type='prob')))
     # Create spatial dataframe for plotting
     df_prob_presence <- as(r_pred[[2]], "SpatialPixelsDataFrame") %>% 
            as.data.frame()
     
  # Return probability of presence raster
  return(df_prob_presence)
  }

# Generate sdm for current climate vars
curr_plot <- predict_dist(curr_clim_stacked, model_fit) %>% 
  mutate(plot_type = "2020 range")

# Generate sdm under future climate vars
fut_plot <- predict_dist(fut_clim_stacked, model_fit) %>% 
  mutate(plot_type = "2050 range")

# Combine all data
all_plots <- rbind(curr_plot,fut_plot)

# Plot scenarios
all_plots %>% 
ggplot() + 
  geom_tile(aes(x=x, y=y, fill=presence)) +
  coord_equal() +
  scale_fill_viridis() +
  theme_bw() +
  facet_wrap(~plot_type)
```

We see the 2020 range is laregely confied to south of the Lake District. A future climate change scenario does not appear significantly different in that there is no significant range expansion north. 

